---
title: "Final Draft"
format: 
  pdf:
    fontsize: 10pt
    include-in-header: 
      text: |
        \usepackage{float}
---

## Introduction
Obesity is a significant public health concern due to its association with various long-term health conditions and increased pressure on healthcare systems. The Scottish Health Survey monitors the health of individuals living in private households in Scotland and provides data that can be used to examine trends in obesity over time. This analysis uses data from the 2008-2012 surveys, including information on age, sex, education, and BMI classification. A binary obesity variable was created from the BMIgroup variable. The aim of this report is to investigate whether obesity prevalence has changed across survey years and whether it differs by demographic characteristics such as age, gender, and education. 
## Exploratory analysis

Prior to performing any analysis, we will first look at our Age variable.

```{r}
#| echo: false 
#| warning: false
#| message: false 

library(tidyverse)
library(ggplot2)
library(sjPlot)
library(broom)
library(performance)
library(yardstick)
library(skimr)
library(janitor)
library(kableExtra) 
library(nnet) 

data<-read.csv("DAProject10.csv") 


data<-data%>% 
  mutate(Obese=as.integer(BMIgroup=="Obese"))%>%
  mutate(Obese=as.factor(Obese), Year=as.factor(Year),
         Sex=as.factor(Sex), Education=as.factor(Education)) 

data$AgeGroup<-cut(data$Age, breaks=c(0,37,50,64,120),
                   labels=c("0-37","38-50","51-64", "65+"))

data<-data%>%
  mutate(AgeGroup=as.factor(AgeGroup))

```

```{r}
#| label: tbl-table_1
#| tbl-cap: "Table of distribution for age variable"  
#| echo: false 

age.summary<-summary(data$Age)
age_table<-as.data.frame(t(as.matrix(age.summary))) 
kable(age_table)
```

We will split our age variable into 4 groups based on the quartiles displayed in @tbl-table_1. Hence, our age groups are as follows: 0-37, 38-50, 51-64 and 65+.

The plots below are barplots of the proportion of obese participants grouped by different categorical variables: Year, Sex, Age Group and Education.

::: {layout-ncol="2" fig-pos="H" fig-height="3" fig-width="4"}
```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by year" 
#| label: fig-1

ggplot(data, aes(x = Obese, group = Year)) +
    geom_bar(aes(y = after_stat(prop), fill = Year), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Year",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by year")

```

```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Sex" 
#| label: fig-2 

ggplot(data, aes(x = Obese, group = Sex)) +
    geom_bar(aes(y = after_stat(prop), fill = Sex), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Sex",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by Sex")

```

```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Age Group" 
#| label: fig-3

ggplot(data, aes(x = Obese, group = AgeGroup)) +
    geom_bar(aes(y = after_stat(prop), fill = AgeGroup), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Age Group",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by age group")
```

```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Education" 
#| label: fig-4 

ggplot(data, aes(x = Obese, group = Education)) +
    geom_bar(aes(y = after_stat(prop), fill = Education), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Education",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by Education")
```
:::

From @fig-1, we can see that the proportion of obese participants are similar for each year of the Scottish Health Study, however, in 2010, the proportion of obese participants is slightly larger than the other years. It appears that the proportion of obese participants is around 30% for all years. Similar can be said regarding @fig-2 , the proportion of obese participants is very similar for both males and females, however is marginally higher for female participants. @fig-3 implies that there is a difference between the proportion of obese participants across age groups, participants of ages 51-64 have the highest obesity proportion (around 35%), and those of ages 0-37 have the lowest (around 20%). @fig-4 also indicates a difference in obesity proportions for those of different education levels. Those who have no qualifications have the highest proportion of obesity (around 36%), and those with a degree or higher have the lowest (around 25%).

From our exploratory analysis, we suspect that the prevalence of obesity is similar over the given years. As well as this, our exploratory analysis indicates there are some differences in obesity by age and socio-economic status or lifestyle factors, however there does not appear to be any major differences in obesity based on gender. We will investigate this further in our formal analysis.

## Formal Analysis

### Question 1 

To assess whether obesity prevalence changed across survey years, a logistic regression model was fitted with obesity status as the binary response variable and year as the explanatory variable, using 2008 as the baseline category. This model is appropriate because the response variable takes two possible values: obese or not obese.

```{r}
#| tbl-cap: "Model summary with 2008 as baseline"
#| label: tbl-table_6
#| echo: false  
#| message: false 
#| fig-pos: "H"

fit_logit_only_year <- glm(Obese ~ Year,
                 data = data,
                 family = binomial)


# Table of model summary with 2008 as a baseline
fit_logit_only_year %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# All Year p-values greated than 0.05, so none are significant

```

**Fitted Formula:**

$$
\ln(\frac{p}{1-p}) = -0.87 + -0.04\cdot \mathbb{I}(\text{Year}=2009) + 0.04\cdot \mathbb I(\text{Year}=2010) + 0.00\cdot \mathbb I(\text{Year}=2011) + 0.00\cdot \mathbb I(\text{Year}=2012)
$$

The fitted model estimates the log-odds of obesity in each year relative to 2008. The intercept of -0.87 represents the log-odds of obesity in 2008. The coefficient for 2009 is -0.04, indicating slightly lower log-odds of obesity than in 2008, while the coefficient for 2010 is 0.04, indicating slightly higher log-odds. The coefficients for 2011 and 2012 are both approximately 0.00, suggesting almost no difference relative to 2008. However, all p-values are greater than 0.05, so none of these year effects are statistically significant. Therefore, there is no evidence that obesity prevalence differed significantly across the survey years considered. ::: {layout-ncol="2" fig-pos="H" fig-height="3" fig-width="4"}

::: {layout-ncol="2" fig-pos="H" fig-height="3" fig-width="4"}
```{r}
#| echo: false 
#| fig-cap: "Point Estimates of log-odds" 
#| label: fig-5 
#| message: false 
#| warning: false 

plot_model(fit_logit_only_year, show.values = TRUE, transform = NULL,
           title = "Log-Odds (Obese)", show.p = TRUE)

```

```{r}
#| echo: false 
#| fig-cap: "Prob of being obese by year with CIs" 
#| label: fig-6
#| message: false 
#| warning: false 
#| fig-align: center

plot_model(fit_logit_only_year, 
           type = "pred", 
           title = "", 
           terms="Year", 
           axis.title = c("Year", "Prob. of being obese"))

```
:::

@fig-5 displays the estimated log-odds of obesity for each year relative to the baseline year. The estimates are all close to zero and their confidence intervals overlap considerably, supporting the conclusion that there are no meaningful differences between years. @fig-6 shows the predicted probability of obesity for each survey year. The predicted probabilities remain relatively stable across years, ranging from approximately 28% to 31%, which is consistent with the exploratory analysis.

```{r}
#| echo: false 
#| fig-cap: "Model Checks" 
#| label: fig-7
#| message: false 
#| warning: false 
#| fig-align: center
#| fig-height: 6
#| fig-width: 8
#| fig-pos: "H"

check_model(fit_logit_only_year, check = c("pp_check","binned_residuals","outliers","qq"))
```
Model diagnostic plots were used to assess the adequacy of the logistic regression model. The posterior predictive check compares observed and predicted counts and suggests that the model reproduces the data reasonably well. The influence plot indicates that no observations exceed Cook’s distance thresholds, suggesting there are no influential data points affecting the model estimates. Overall, the diagnostic plots do not indicate any major issues with model fit.

### Question 2
