---
title: "An investigation into various factors that impact obesity in Scotland from 2008-2012"
format: 
  pdf:
    embed-resources: true
    fontsize: 9pt
    include-in-header: 
      text: |
        \usepackage{float}
---


# 1. Introduction

This project analyses data from the Scottish Health Survey (2008–2012) to investigate obesity prevalence 
and its association with demographic and socioeconomic factors.

The main research questions are:

1. Has obesity prevalence changed between 2008 and 2012?
2. Is obesity associated with age, sex, and education level?


## Exploratory analysis

Prior to performing any analysis, we will first look at our Age variable.

```{r}
#| echo: false 
#| warning: false
#| message: false 

library(tidyverse)
library(ggplot2)
library(sjPlot)
library(broom)
library(performance)
library(yardstick)
library(skimr)
library(janitor)
library(kableExtra) 
library(nnet) 

data<-read.csv("DAProject10.csv") 


data<-data%>% 
  mutate(Obese=as.integer(BMIgroup=="Obese"))%>%
  mutate(Obese=as.factor(Obese), Year=as.factor(Year),
         Sex=as.factor(Sex), Education=as.factor(Education)) 

data$AgeGroup<-cut(data$Age, breaks=c(0,37,50,64,120),
                   labels=c("0-37","38-50","51-64", "65+"))

data<-data%>%
  mutate(AgeGroup=as.factor(AgeGroup))

```

```{r}
#| label: tbl-table_1
#| tbl-cap: "Table of distribution for age variable"  
#| echo: false 

age.summary<-summary(data$Age)
age_table<-as.data.frame(t(as.matrix(age.summary))) 
kable(age_table)
```

We will split our age variable into 4 groups based on the quartiles displayed in @tbl-table_1. Hence, our age groups are as follows: 0-37, 38-50, 51-64 and 65+.

The plots below are barplots of the proportion of obese participants grouped by different categorical variables: Year, Sex, Age Group and Education. 

From @fig-1, we can see that the proportion of obese participants are similar for each year of the Scottish Health Study, however, in 2010, the proportion of obese participants is slightly larger than the other years. It appears that the proportion of obese participants is around 30% for all years. Similar can be said regarding @fig-2 , the proportion of obese participants is very similar for both males and females, however is marginally higher for female participants. @fig-3 implies that there is a difference between the proportion of obese participants across age groups, participants of ages 51-64 have the highest obesity proportion (around 35%), and those of ages 0-37 have the lowest (around 20%). @fig-4 also indicates a difference in obesity proportions for those of different education levels. Those who have no qualifications have the highest proportion of obesity (around 36%), and those with a degree or higher have the lowest (around 25%).

::: {layout-ncol="2" fig-pos="H" fig-height="3" fig-width="4"}
```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by year" 
#| label: fig-1

ggplot(data, aes(x = Obese, group = Year)) +
    geom_bar(aes(y = after_stat(prop), fill = Year), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Year",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by year")

```

```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Sex" 
#| label: fig-2 

ggplot(data, aes(x = Obese, group = Sex)) +
    geom_bar(aes(y = after_stat(prop), fill = Sex), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Sex",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by Sex")

```

```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Age Group" 
#| label: fig-3

ggplot(data, aes(x = Obese, group = AgeGroup)) +
    geom_bar(aes(y = after_stat(prop), fill = AgeGroup), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Age Group",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by age group")
```

```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Education" 
#| label: fig-4 

ggplot(data, aes(x = Obese, group = Education)) +
    geom_bar(aes(y = after_stat(prop), fill = Education), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Education",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by Education")
```
:::



From our exploratory analysis, we suspect that the prevalence of obesity is similar over the given years. As well as this, our exploratory analysis indicates there are some differences in obesity by age and socio-economic status or lifestyle factors, however there does not appear to be any major differences in obesity based on gender. We will investigate this further in our formal analysis.


## Formal Analysis 

### Investigating changes in prevalence of obesity in Scotland over the years 2008-2012 

```{r}
#| tbl-cap: "Model summary with 2008 as baseline"
#| label: tbl-table_2
#| echo: false  
#| message: false 
#| fig-pos: "H"

fit_logit_only_year <- glm(Obese ~ Year,
                 data = data,
                 family = binomial)

# Table of model summary with 2008 as a baseline
fit_logit_only_year %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# All Year p-values greated than 0.05, so none are significant 

``` 
$$
\ln(\frac{p}{1-p}) = -0.87 + -0.04\cdot \mathbb{I}(\text{Year}=2009) + 0.04\cdot \mathbb I(\text{Year}=2010) + 0.00\cdot \mathbb I(\text{Year}=2011) + 0.00\cdot \mathbb I(\text{Year}=2012)
$$ 

::: {layout-ncol="2" fig-pos="H" fig-height="3" fig-width="4"}
```{r}
#| echo: false 
#| fig-cap: "Point Estimates of log-odds"
#| label: fig-5 
#| message: false 
#| warning: false 

# Point estimates of the log-odds
plot_model(fit_logit_only_year, show.values = TRUE, transform = NULL,
           title = "Log-Odds (Obese)", show.p = TRUE) 

```

```{r}
#| echo: false 
#| fig-cap: "Prob of being obese by year with CIs" 
#| label: fig-6
#| message: false 
#| warning: false 
#| fig-align: center

# Probability of being obese by year with confidence intervals
plot_model(fit_logit_only_year, 
           type = "pred", 
           title = "", 
           terms="Year", 
           axis.title = c("Year", "Prob. of being obese"))
``` 
:::

```{r}
#| echo: false 
#| fig-cap: "ROC curve" 
#| label: fig-7
#| message: false 
#| warning: false 
#| fig-align: center 
#| fig-height: 2
#| fig-width: 2 
#| fig-pos: "H"

pred_results1 <- fit_logit_only_year %>% 
  broom::augment(type.predict = "response") %>% 
  mutate(predicted_class = factor(ifelse(.fitted > 0.5, "Obese", "Not Obese"), levels = levels(Obese))) 

roc_curve(pred_results1,
          truth=Obese, 
          .fitted,
          event_level="second")%>%
  autoplot()
```

### Investigating the differences in obesity by age, gender, socio-economic status and lifestyle factors 

$$
y_i\sim \text{Bernoulli}(p_i)\\\text{logit}(p_i)=\bf{x}_i^\top\boldsymbol\beta, \,\,\,i=
{1,...,25224}
$$ where $\bf x^\top_i$ is a vector containing predictor variables and $\boldsymbol\beta$ is a vector of coefficient estimates. So that, 

$$
\bf{x}_i=\begin{pmatrix}1\\x_{35-50}\\x_{51-64}\\x_{65+}\\x_{\text{edu=higher}}\\⋮\\x_{\text{edu=standard}}\\x_{\text{male}}\end{pmatrix}\,\,\,\,\boldsymbol\beta=\begin{pmatrix}\alpha\\\beta_{38-50}\\\beta{51-64}\\\beta_{65+}\\\beta_{\text{edu=higher}}\\⋮\\\beta_{\text{edu=standard}}\\\beta_{\text{male}}\end{pmatrix}
$$  

where the reference categories for age group, education and sex are 0-37, degree or higher and female, respectively. @tbl-table_4 contains the details of the fitted model.

```{r}
#| tbl-cap: "Model summary of Obesity with Agegroup, Sex and Education"
#| label: tbl-table_4
#| echo: false  
#| message: false 
#| fig-pos: "H"

#Logistic regression model with all variables 

fit_logit_q2 <- glm(
  Obese ~ AgeGroup + Education + Sex,
  data = data,
  family = binomial
)

#Table of Full model 

tab_model(
  fit_logit_q2,
  show.ci = 0.95,
  show.aic = TRUE,
  transform = "exp"
)


``` 

```{r}
#| tbl-cap: "Table comparing AIC values of null model, full model and reduced model (without Sex)"
#| label: tbl-table_6
#| echo: false  
#| message: false 
#| fig-pos: "H" 

# Null model (intercept only, no predictors)
fit_logit_q2_null <- glm(
  Obese ~ 1,
  data = data,
  family = binomial
)

#Model with sex removed
fit_logit_q2_nosex <- glm(
  Obese ~ AgeGroup + Education,
  data = data,
  family = binomial
)

AICs<-AIC(
  fit_logit_q2_null,
  fit_logit_q2,
  fit_logit_q2_nosex
  
)
kable(AICs) 

```




```{r}
#| tbl-cap: "Table of model with sex removed"
#| label: tbl-table_7
#| echo: false  
#| message: false 
#| fig-pos: "H"  

tab_model(
  fit_logit_q2_nosex,
  show.ci = 0.95,
  show.aic = TRUE,
  transform = "exp"
)

```


```{r}
#| echo: false 
#| fig-cap: "Plot of odds of chosen model" 
#| label: fig-8
#| message: false 
#| warning: false 
#| fig-align: center 
#| fig-height: 3
#| fig-width: 4
#| fig-pos: "H"

plot_model(fit_logit_q2_nosex, show.values = TRUE)

``` 

## Conclusion










































