---
title: "Project 10 – Obesity Prevalence"
author: Jasmine
format: html
---

# 1. Introduction

This project analyses data from the Scottish Health Survey (2008–2012) to investigate obesity prevalence 
and its association with demographic and socioeconomic factors.

The main research questions are:

1. Has obesity prevalence changed between 2008 and 2012?
2. Is obesity associated with age, sex, and education level?

```{r}
library(tidyverse)

data <- read.csv("~/Desktop/data/DAProject10.csv")

str(data)
summary(data)
```
```{r}
table(data$BMIgroup)
prop.table(table(data$BMIgroup))
```


# Convert categorical variables to factors
# This ensures that logistic regression treats them as categorical predictors
# and sets appropriate reference levels.
data <- data %>%
  mutate(
    Sex = factor(Sex),
    Education = factor(Education),
    Year = factor(Year),
    BMIgroup = factor(BMIgroup, levels = c("Underweight","Normal","Overweight","Obese"))
  )



# Create a binary obesity variable (1 = Obese, 0 = Not Obese)
# This allows us to compute obesity prevalence by year
obese_by_year <- data %>%
  mutate(obese = (BMIgroup == "Obese")) %>%
  group_by(Year) %>%                     # sample size per year
  summarise(                             # proportion obese
    n = n(),
    obese_rate = mean(obese)
  )

obese_by_year


# Plot obesity prevalence over time
# Year is on the x-axis and obesity rate on the y-axis
# group = 1 ensures points are connected across years
ggplot(obese_by_year, aes(x = Year, y = obese_rate, group = 1)) +
  geom_line() +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Obesity Prevalence by Year",
    y = "Obesity Rate",
    x = "Year"
  )


# Chi-squared test of independence:
# H0: BMI category (BMIgroup) is independent of survey year (Year)
# H1: BMI category distribution differs across years
chisq.test(table(data$BMIgroup, data$Year))


# Create a binary outcome for logistic regression:
# obese = 1 if BMIgroup is "Obese", otherwise 0
data <- data %>%
mutate(obese = as.integer(BMIgroup == "Obese"))


# Logistic regression model:
# Outcome: obesity status (1 = obese)
# Predictors: Age, Sex, Education
# Binomial family with logit link models log-odds of obesity
fit_logit <- glm(obese ~ Age + Sex + Education,
                 data = data,
                 family = binomial)
summary(fit_logit)
exp(cbind(OR = coef(fit_logit), confint(fit_logit)))


fit_logit2 <- glm(obese ~ Age + Sex + Education + Year,
                  data = data,
                  family = binomial)
fit_logit2


# Fit logistic regression model without Year
# This model assesses whether Age, Sex, and Education
# are associated with obesity
fit_logit_without_year <- glm(
  obese ~ Age + Sex + Education,
  data = data,
  family = binomial
)

# Fit logistic regression model including Year
# This model tests whether Year provides additional
# explanatory power beyond demographic variables
fit_logit_with_year <- glm(
  obese ~ Age + Sex + Education + Year,
  data = data,
  family = binomial
)

# Compare nested models using Likelihood Ratio Test
# to assess whether adding Year significantly improves model fit
anova(fit_logit_without_year, fit_logit_with_year, test="Chisq")














































