---
title: "An investigation into various factors that impact obesity in Scotland from 2008-2012"
format: 
  pdf:
    embed-resources: true
    fontsize: 9pt
    include-in-header: 
      text: |
        \usepackage{float}
---


# 1. Introduction

This project analyses data from the Scottish Health Survey (2008–2012) to investigate obesity prevalence 
and its association with demographic and socioeconomic factors.

The main research questions are:

1. Has obesity prevalence changed between 2008 and 2012?
2. Is obesity associated with age, sex, and education level?


## Exploratory analysis

Prior to performing any analysis, we will first look at our Age variable.

```{r}
#| echo: false 
#| warning: false
#| message: false 

library(tidyverse)
library(ggplot2)
library(sjPlot)
library(broom)
library(performance)
library(yardstick)
library(skimr)
library(janitor)
library(kableExtra) 
library(nnet) 

data<-read.csv("DAProject10.csv") 


data<-data%>% 
  mutate(Obese=as.integer(BMIgroup=="Obese"))%>%
  mutate(Obese=as.factor(Obese), Year=as.factor(Year),
         Sex=as.factor(Sex), Education=as.factor(Education)) 

data$AgeGroup<-cut(data$Age, breaks=c(0,37,50,64,120),
                   labels=c("0-37","38-50","51-64", "65+"))

data<-data%>%
  mutate(AgeGroup=as.factor(AgeGroup))

```

```{r}
#| label: tbl-table_1
#| tbl-cap: "Table of distribution for age variable"  
#| echo: false 

age.summary<-summary(data$Age)
age_table<-as.data.frame(t(as.matrix(age.summary))) 
kable(age_table)
```

We will split our age variable into 4 groups based on the quartiles displayed in @tbl-table_1. Hence, our age groups are as follows: 0-37, 38-50, 51-64 and 65+.

The plots below are barplots of the proportion of obese participants grouped by different categorical variables: Year, Sex, Age Group and Education. 

From @fig-1, we can see that the proportion of obese participants are similar for each year of the Scottish Health Study, however, in 2010, the proportion of obese participants is slightly larger than the other years. It appears that the proportion of obese participants is around 30% for all years. Similar can be said regarding @fig-2 , the proportion of obese participants is very similar for both males and females, however is marginally higher for female participants. @fig-3 implies that there is a difference between the proportion of obese participants across age groups, participants of ages 51-64 have the highest obesity proportion (around 35%), and those of ages 0-37 have the lowest (around 20%). @fig-4 also indicates a difference in obesity proportions for those of different education levels. Those who have no qualifications have the highest proportion of obesity (around 36%), and those with a degree or higher have the lowest (around 25%).

::: {layout-ncol="2" fig-pos="H" fig-height="3" fig-width="4"}
```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by year" 
#| label: fig-1

ggplot(data, aes(x = Obese, group = Year)) +
    geom_bar(aes(y = after_stat(prop), fill = Year), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Year",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by year")

```

```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Sex" 
#| label: fig-2 

ggplot(data, aes(x = Obese, group = Sex)) +
    geom_bar(aes(y = after_stat(prop), fill = Sex), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Sex",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by Sex")

```

```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Age Group" 
#| label: fig-3

ggplot(data, aes(x = Obese, group = AgeGroup)) +
    geom_bar(aes(y = after_stat(prop), fill = AgeGroup), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Age Group",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by age group")
```

```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Education" 
#| label: fig-4 

ggplot(data, aes(x = Obese, group = Education)) +
    geom_bar(aes(y = after_stat(prop), fill = Education), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Education",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by Education")
```
:::



From our exploratory analysis, we suspect that the prevalence of obesity is similar over the given years. As well as this, our exploratory analysis indicates there are some differences in obesity by age and socio-economic status or lifestyle factors, however there does not appear to be any major differences in obesity based on gender. We will investigate this further in our formal analysis.


## Formal Analysis 

### Investigating changes in prevalence of obesity in Scotland over the years 2008-2012 

```{r}
#| tbl-cap: "Model summary with 2008 as baseline"
#| label: tbl-table_2
#| echo: false  
#| message: false 
#| fig-pos: "H"

fit_logit_only_year <- glm(Obese ~ Year,
                 data = data,
                 family = binomial)

# Table of model summary with 2008 as a baseline
fit_logit_only_year %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# All Year p-values greated than 0.05, so none are significant 

``` 
$$
\ln(\frac{p}{1-p}) = -0.87 + -0.04\cdot \mathbb{I}(\text{Year}=2009) + 0.04\cdot \mathbb I(\text{Year}=2010) + 0.00\cdot \mathbb I(\text{Year}=2011) + 0.00\cdot \mathbb I(\text{Year}=2012)
$$ 

::: {layout-ncol="2" fig-pos="H" fig-height="3" fig-width="4"}
```{r}
#| echo: false 
#| fig-cap: "Point Estimates of log-odds"
#| label: fig-5 
#| message: false 
#| warning: false 

# Point estimates of the log-odds
plot_model(fit_logit_only_year, show.values = TRUE, transform = NULL,
           title = "Log-Odds (Obese)", show.p = TRUE) 

```

```{r}
#| echo: false 
#| fig-cap: "Prob of being obese by year with CIs" 
#| label: fig-6
#| message: false 
#| warning: false 
#| fig-align: center

# Probability of being obese by year with confidence intervals
plot_model(fit_logit_only_year, 
           type = "pred", 
           title = "", 
           terms="Year", 
           axis.title = c("Year", "Prob. of being obese"))
``` 
:::

```{r}
#| echo: false 
#| fig-cap: "ROC curve" 
#| label: fig-7
#| message: false 
#| warning: false 
#| fig-align: center 
#| fig-height: 2
#| fig-width: 2 
#| fig-pos: "H"

pred_results1 <- fit_logit_only_year %>% 
  broom::augment(type.predict = "response") %>% 
  mutate(predicted_class = factor(ifelse(.fitted > 0.5, "Obese", "Not Obese"), levels = levels(Obese))) 

roc_curve(pred_results1,
          truth=Obese, 
          .fitted,
          event_level="second")%>%
  autoplot()
```


Posterior predictive check (Top-left):
Compares the observed data (number of obese vs not obese) against the predicted number
using simulated data. The values appear identical to predictions.

Binned residuals (Top-right):
Bin the observations based on their fitted values, and average the residual
value in each bin. The plots shows the average residual values versus the average
fitted value for each bin. All residuals fall outside the error bounds

Influential observations (Bottom-left):
This plot is used to identify influential observations. If any points in this plot
fall outside of Cook’s distance (the dashed lines) then it is considered an influential
observation. All values fall within the dashed lines.

Uniformity of residuals (Bottom-right):
Since our response is not normally distributed, residuals are not expected to be
normally distributed either. Thus, we can use QQ plots to check the uniformity of
residuals. The values fall exactly along the line.


```{r}
# Logistic regression model:
# Outcome: obesity status (1 = Obese)
# Predictors: Age, Sex, Education
# Binomial family with logit link models log-odds of obesity


# Since all terms are categorical, fit the full model with multinomial, family = binomial
model_full <- multinom(Obese ~ AgeGroup + Education + Sex, data = data, family = binomial)

# Table of full model summary
model_full %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# Sex has p-value>0.05, so is removed

# Fit the final model, without Sex
model_final <- multinom(Obese ~ AgeGroup + Education, data = data, family = binomial)

# Table of final model summary
model_final %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# All p-values<0.05, so the final model is good

# Compare performance of model_final and model_full
compare_performance(model_final,model_full,metrics = "AIC")
# model_final has lower AIC

# Log-odds for model_final
plot_model(model_final, show.values = TRUE, transform = NULL,
           title = "Log-Odds (Obese)", show.p = TRUE)

check_model




# # Fit logistic regression model without Year
# # This model assesses whether Age, Sex, and Education
# # are associated with obesity
# fit_logit_without_year <- glm(
#   Obese ~ AgeGroup + Sex + Education,
#   data = data,
#   family = binomial
# )


# # Fit logistic regression model including Year
# # This model tests whether Year provides additional
# # explanatory power beyond demographic variables
# fit_logit_with_year <- glm(
#   Obese ~ AgeGroup + Sex + Education + Year,
#   data = data,
#   family = binomial
# )

# Compare nested models using Likelihood Ratio Test
# to assess whether adding Year significantly improves model fit
anova(fit_logit_without_year, fit_logit_with_year, test="Chisq")


# Compare AIC, fit_logit_without_year is lowest
AIC(fit_logit_without_year, fit_logit_with_year, fit_logit_only_year)



# Table of model summary
fit_logit_without_year %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# Sex[Male] has p-value 0.328, so is not significant


# Fit final model without Year and Sex
fit_logit_final <- glm(
  Obese ~ AgeGroup + Education,
  data = data,
  family = binomial
)

# Table of final model summary
fit_logit_final %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# All p-values are less than 0.05, so all are significant



plot_model(type = "pred",
           fit_logit_final,
           terms = "Education",
           axis.title = c("Education Category",
                          "Prob. of being obese"),
           title = "Plot of Education coefficients")

plot_model(type = "pred",
           fit_logit_final,
           terms = "AgeGroup",
           axis.title = c("Age Group",
                          "Prob. of being obese"),
           title = "Plot of AgeGroup coefficients")

plot_model(type = "pred",
           fit_logit_final,
           terms = c("AgeGroup", "Education"),
           axis.title = c("Age Group",
                          "Prob. of being obese", "Education"),
           title = "Plot of AgeGroup and Education coefficients")


# Point estimates of the log-odds for the final model with AgeGroup and Education
plot_model(fit_logit_final, show.values = TRUE, transform = NULL,
           title = "Log-Odds (Obese)", show.p = TRUE)


check_model(fit_logit_final, check = c("pp_check","binned_residuals","outliers","qq"))
# Binned residuals plot has most values outside error bounds
# All other plots satisfy the assumptions

```





















1111






















