---
title: "Project 10 – Obesity Prevalence"
author: Jasmine
format: html
---

# 1. Introduction

This project analyses data from the Scottish Health Survey (2008–2012) to investigate obesity prevalence 
and its association with demographic and socioeconomic factors.

The main research questions are:

1. Has obesity prevalence changed between 2008 and 2012?
2. Is obesity associated with age, sex, and education level?

```{r}
#| echo: false 
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(sjPlot)
library(broom)
library(performance)
library(yardstick)
library(skimr) 
library(janitor)
library(kableExtra)

data <- read.csv("DAProject10.csv")  # Load dataset from project directory

```



```{r}
table(data$BMIgroup)
prop.table(table(data$BMIgroup))



# Convert categorical variables to factors
# This ensures that logistic regression treats them as categorical predictors
# and sets appropriate reference levels.
data <- data %>%
  mutate(
    Sex = factor(Sex),
    Education = factor(Education),
    Year = factor(Year),
    BMIgroup = factor(BMIgroup, levels = c("Underweight","Normal","Overweight","Obese"))
  )

# Converts age into categorical and make it a factor
data$AgeGroup<-cut(data$Age, breaks=c(0,37,50,64,120),
                   labels=c("0-37", "38-50", "51-64", "65+"))
data <- data %>%
  mutate(AgeGroup = as.factor(AgeGroup))

# Create a binary outcome for logistic regression:
# Obese = 1 if BMIgroup is "Obese", otherwise 0, then converts to a factor
data<-data %>%
  mutate(Obese=as.integer(BMIgroup=="Obese"))%>%
  mutate(Obese=as.factor(Obese))


# # Create a binary obesity variable (1 = Obese, 0 = Not Obese)
# # This allows us to compute obesity prevalence by year
# obese_by_year <- data %>%
#   mutate(Obese = (BMIgroup == "Obese")) %>%
#   group_by(Year) %>%                     # sample size per year
#   summarise(                             # proportion obese
#     n = n(),
#     obese_rate = mean(Obese),
#     obese_num = obese_rate*n
#   )
# 
# obese_by_year


# # Plot obesity prevalence over time
# # Year is on the x-axis and obesity rate on the y-axis
# # group = 1 ensures points are connected across years
# ggplot(obese_by_year, aes(x = Year, y = obese_rate, group = 1)) +
#   geom_line() +
#   geom_point(size = 2) +
#   scale_y_continuous(labels = scales::percent_format()) +
#   labs(
#     title = "Obesity Prevalence by Year",
#     y = "Obesity Rate",
#     x = "Year"
#   )


# # Chi-squared test of independence:
# # H0: BMI category (BMIgroup) is independent of survey year (Year)
# # H1: BMI category distribution differs across years
# chisq.test(table(data$BMIgroup, data$Year))



# # Bar plot of obesity prevalence over time
# # Year is on the x-axis and obesity rate on the y-axis
# ggplot(data, aes(x=Year, fill=Obese))+
#   geom_bar()

# Bar plot of obesity proportion by year
ggplot(data, aes(x = Obese, group = Year)) +
  geom_bar(aes(y = after_stat(prop), fill = Year), 
           stat = "count", position = "dodge") +
  labs(y = "Proportion", fill = "Year",
       x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by year")


# Q1 logistic regression model:
fit_logit_only_year <- glm(Obese ~ Year,
                 data = data,
                 family = binomial)

# Table of model summary with 2008 as a baseline
fit_logit_only_year %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# All Year p-values greated than 0.05, so none are significant

# Model formula
# ln(p/1-p) = -0.87 + -0.04*I(Year=2009) + 0.04*I(Year=2010)

# Point estimates of the log-odds
plot_model(fit_logit_only_year, show.values = TRUE, transform = NULL,
           title = "Log-Odds (Obese)", show.p = TRUE)

# Probability of being obese by year with confidence intervals
plot_model(fit_logit_only_year, 
           type = "pred", 
           title = "", 
           terms="Year", 
           axis.title = c("Year", "Prob. of being obese"))


check_model(fit_logit_only_year, check = c("pp_check","binned_residuals","outliers","qq"))

```

Posterior predictive check (Top-left):
Compares the observed data (number of obese vs not obese) against the predicted number
using simulated data. The values appear identical to predictions.

Binned residuals (Top-right):
Bin the observations based on their fitted values, and average the residual
value in each bin. The plots shows the average residual values versus the average
fitted value for each bin. All residuals fall outside the error bounds

Influential observations (Bottom-left):
This plot is used to identify influential observations. If any points in this plot
fall outside of Cook’s distance (the dashed lines) then it is considered an influential
observation. All values fall within the dashed lines.

Uniformity of residuals (Bottom-right):
Since our response is not normally distributed, residuals are not expected to be
normally distributed either. Thus, we can use QQ plots to check the uniformity of
residuals. The values fall exactly along the line.


```{r}
# Logistic regression model:
# Outcome: obesity status (1 = Obese)
# Predictors: Age, Sex, Education
# Binomial family with logit link models log-odds of obesity



# Fit logistic regression model without Year
# This model assesses whether Age, Sex, and Education
# are associated with obesity
fit_logit_without_year <- glm(
  Obese ~ AgeGroup + Sex + Education,
  data = data,
  family = binomial
)

# Fit logistic regression model including Year
# This model tests whether Year provides additional
# explanatory power beyond demographic variables
fit_logit_with_year <- glm(
  Obese ~ AgeGroup + Sex + Education + Year,
  data = data,
  family = binomial
)

# Compare nested models using Likelihood Ratio Test
# to assess whether adding Year significantly improves model fit
anova(fit_logit_without_year, fit_logit_with_year, test="Chisq")


# Compare AIC, fit_logit_without_year is lowest
AIC(fit_logit_without_year, fit_logit_with_year, fit_logit_only_year)

# Table of model summary
fit_logit_with_year %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)

# Table of model summary
fit_logit_without_year %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# Sex[Male] has p-value 0.328, so is not significant


# Fit final model without Year and Sex
fit_logit_final <- glm(
  Obese ~ AgeGroup + Education,
  data = data,
  family = binomial
)

# Table of final model summary
fit_logit_final %>% tab_model(show.ci = 0.95, show.aic=TRUE, transform = NULL)
# All p-values are less than 0.05, so all are significant



plot_model(type = "pred",
           fit_logit_final,
           terms = "Education",
           axis.title = c("Education Category",
                          "Prob. of being obese"),
           title = "Plot of Education coefficients")

plot_model(type = "pred",
           fit_logit_final,
           terms = "AgeGroup",
           axis.title = c("Age Group",
                          "Prob. of being obese"),
           title = "Plot of AgeGroup coefficients")

plot_model(type = "pred",
           fit_logit_final,
           terms = c("AgeGroup", "Education"),
           axis.title = c("Age Group",
                          "Prob. of being obese", "Education"),
           title = "Plot of AgeGroup and Education coefficients")

```





















1111






















