---
title: "lab draft"
format: html
editor: visual
---

##Intro 

## Exploratory Analysis

Prior to performing any analysis, we will first look at our Age variable.

```{r}
#| echo: false 
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(sjPlot)
library(broom)
library(performance)
library(yardstick)
library(skimr)
library(janitor)
library(kableExtra) 
library(nnet)

data<-read.csv("DAProject10.csv")


data<-data%>% 
  mutate(Obese=as.integer(BMIgroup=="Obese"))%>%
  mutate(Obese=as.factor(Obese), Year=as.factor(Year),
         Sex=as.factor(Sex), Education=as.factor(Education))

data$AgeGroup<-cut(data$Age, breaks=c(0,37,50,64,120),
                   labels=c("0-37","38-50","51-64", "65+"))

data<-data%>%
  mutate(AgeGroup=as.factor(AgeGroup))
```

```{r}
#| echo: false 
#| label: tbl-table_1
#| tbl-cap: "Table of distribution for age variable" 

age.summary<-summary(data$Age)
age.summary

```

We will split our age variable into 4 groups based on the quartiles displayed in @tbl-table_1. Hence, our age groups are as follows: 0-37, 38-50, 51-64 and 65+.

@tbl-table_2 displays the total number of participants in the Scottish Health Surveys for each of the years 2008-2012, as well as the number of obese participants. It also includes the proportion of participants who are obese and not obese. From this, it can be seen that the year with the largest proportion of obese participants is 2010 (30.47%), and year with the smallest proportion of obese participants is 2009 (28.79%), though there is not a large difference across the years.

```{r}
#| tbl-cap: "Summary Statistics Q1"
#| label: tbl-table_2 
#| echo: false 


summary1<-data%>%
  summarise(n=n(),
            obese_n=sum(Obese==1),
            obese.prop=mean(Obese==1),
            not.obese.prop=mean(Obese==0),
            .by=Year) 
kable(summary1, digits=4, col.names=c("Year", "Number of participants (total)", "Number of participants (obese)", "Proportion of participants (obese)", "Proportion of participants (not obese)"))

```

@tbl-table_3 displays the total number of participants in the Scottish Health Surveys for the years 2008-2012, grouped by sex, as well as the number of obese participants. It also includes the proportion of participants who are obese and not obese. From this, it can be seen that females have a slightly higher proportion of obese participants when compared to male participants (29.81% vs. 29.37%.)

```{r}
#| tbl-cap: "Summary Statistics Q2 (sex)"
#| label: tbl-table_3 
#| echo: false

summary2sex<-data%>%
  summarise(n=n(),
            obese_n=sum(Obese==1),
            obese.prop=mean(Obese==1),
            not.obese.prop=mean(Obese==0),
            .by=Sex)  
summary2age<-data%>%
  summarise(n=n(),
            obese_n=sum(Obese==1),
            obese.prop=mean(Obese==1),
            not.obese.prop=mean(Obese==0),
            .by=AgeGroup) 
summary2Ed<-data%>%
  summarise(n=n(),
            obese_n=sum(Obese==1),
            obese.prop=mean(Obese==1),
            not.obese.prop=mean(Obese==0),
            .by=Education)

kable(summary2sex, digits=4, col.names=c("Sex", "Number of participants (total)", "Number of participants (obese)", "Proportion of participants (obese)", "Proportion of participants (not obese)"))

```

@tbl-table_4 displays the total number of participants in the Scottish Health Surveys for the years 2008-2012, grouped by age group, as well as the number of obese participants. It also includes the proportion of participants who are obese and not obese. From this, it can be seen that participants of ages 51-64 (35.14%) have the highest proportion of obesity, and those of ages 0-37 have the lowest proportion of obesity (20.36%).

```{r}
#| tbl-cap: "Summary Statistics Q2 (age group)"
#| label: tbl-table_4
#| echo: false 

kable(summary2age, digits=4, col.names=c("Age group", "Number of participants (total)", "Number of participants (obese)", "Proportion of participants (obese)", "Proportion of participants (not obese)"))


```

@tbl-table_5 displays the total number of participants in the Scottish Health Surveys for the years 2008-2012, grouped by education, as well as the number of obese participants. It also includes the proportion of participants who are obese and not obese. From this, it can be seen that those with no qualifications have the highest proportion of obesity (36.47%), whereas those with a level of education of a degree or higher have the lowest proportion (24.77%).

```{r}
#| tbl-cap: "Summary Statistics Q2 (education)"
#| label: tbl-table_5
#| echo: false 

kable(summary2Ed, digits=4, col.names=c("Education", "Number of participants (total)", "Number of participants (obese)", "Proportion of participants (obese)", "Proportion of participants (not obese)"))

```

The plots below are barplots of the proportion of obese participants grouped by different categorical variables: Year, Sex, Age Group and Education.

::::::: columns
::: column
```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by year" 
#| label: fig-1

ggplot(data, aes(x = Obese, group = Year)) +
    geom_bar(aes(y = after_stat(prop), fill = Year), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Year",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by year")

```
:::

::: column
```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Sex" 
#| label: fig-2 

ggplot(data, aes(x = Obese, group = Sex)) +
    geom_bar(aes(y = after_stat(prop), fill = Sex), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Sex",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by Sex")


```
:::

::: column
```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Age Group" 
#| label: fig-3

ggplot(data, aes(x = Obese, group = AgeGroup)) +
    geom_bar(aes(y = after_stat(prop), fill = AgeGroup), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Age Group",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by age group")
```
:::

::: column
```{r}
#| echo: false 
#| fig-cap: "Barplot of obesity status by Education" 
#| label: fig-4 

ggplot(data, aes(x = Obese, group = Education)) +
    geom_bar(aes(y = after_stat(prop), fill = Education), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Education",
         x="Obese (0-no, 1-yes)")+ 
  ggtitle("Barplot of obesity status by Education")
```
:::
:::::::

From @fig-1, we can see that the proportion of obese participants are similar for each year of the Scottish Health Study, however, in 2010, the proportion of obese participants is slightly larger than the other years. Similar can be said regarding @fig-2 , the proportion of obese participants is very similar for both males and females, however is marginally higher for female participants. @fig-3 implies that there is a difference between the proportion of obese participants across age groups, as seen from the summary statistics in @tbl-table_3, participants of ages 51-64 have the highest obesity proportion, and those of ages 0-37 have the lowest. @fig-4 also indicates a difference in obesity proportions for those of different education levels. Those who have no qualifications have the highest proportion of obesity, and those with a degree or higher have the lowest, as indicated from the summary statistics of @tbl-table_4.

From our exploratory analysis, we suspect that the prevalence of obesity is similar over the given years. As well as this, our exploratory analysis indicates there are some differences in obesity by age and socio-economic status or lifestyle factors, however there does not appear to be any major differences in obesity based on gender. We will investigate this further in our formal analysis.

##Formal Analysis
